Emulation of high multiplicity matrix elements for electron-positron and hadron-hadron colliders

### 2022-11-02 intro.tex
proof read chapter 1

    The SM is a gauge quantum field theory (QFT) where particles
    are described as excitations in quantum fields. It can be specified
    by the gauge group
    \begin{equation}\label{eqn:SM_gauge}
        \mathrm{SU}(3)_{c} \times \mathrm{SU}(2)_{L} \times \mathrm{U}(1)_{Y}\, ,
    \end{equation}
    where subscripts denote the charges of the gauge groups. The first gauge
    group with colour charge $C$ describes the interactions of the
    strong force within the theory of quantum chromodynamics (QCD), which we
    will elaborate more on in the next section. The second
    and third gauge groups represent the electroweak sector of the SM with
    charges $L$ for left and $Y$ for hypercharge. Under electroweak
    spontaneous symmetry breaking (EWSB), this product becomes
    \begin{equation}\label{eqn:SM_SSB}
        \mathrm{SU}(2)_{L} \times \mathrm{U}(1)_{Y} \xrightarrow{\mathrm{EWSB}} \mathrm{U}(1)_{\mathrm{EM}} \, ,
    \end{equation}
    giving rise to the electromagnetic and weak forces we observe.
    The effect of gravity is considered to be negligible on the scales
    considered in the SM and so it is not described.
    Each of the three fundamental forces described by the SM is mediated
    by the exchange of a gauge boson. The gauge bosons mediating the strong
    and EM force are the massless gluon $g$ and massless photon $\gamma$,
    respectively. For the weak force the gauge bosons are the $W^{\pm}$
    and $Z^{0}$ bosons which attain a mass through the Higgs mechanism \cite{Englert:1964et,Higgs:1964pj,Guralnik:1964eu}
    during EWSB, which elucidates the Higgs boson $H$.

@@

\section{Introduction to quantum chromodynamics}\label{sec:qcd}
    \subsection{The QCD Lagrangian}\label{sec:L_qcd}
    Quantum chromodynamics (QCD) is the sector of the SM that
    describes the strong interaction. 
    QCD is a non-Abelian gauge theory with gauge group
    $\mathrm{SU}(3)_{c}$ where the charge is named colour.
    The gauge and fermion part of the QCD Lagrangian is
    \begin{equation}\label{eqn:L_QCD}
        \mathcal{L}_{\mathrm{QCD}} = -\dfrac{1}{4}F^{a}_{\mu\nu}F^{a, \, \mu\nu}
        + \sum_{f} \bar{\psi}_{i}^{f}(\mathrm{i}\slashed{D}_{ij} - \delta_{ij}m_{f})\psi_{j}^{f} \, ,
    \end{equation}
    where twice repeated indices are summed over.
    The fields $\psi_{i}^{f}$ ($\bar{\psi}_{i}^{f}$) are the fermions (antifermions),
    representing quarks (antiquarks) with flavours $f$: 
    up $u$, down $d$, strange $s$, charm $c$, top $t$, and bottom $b$,
    with masses $m_{f}$. They transform under the fundamental
    representation with indices $i, j \in \{1, 2, 3\}$, named colour indices.
    The gauge fields $A^{a}_{\mu}$, corresponding to gluons, 
    appear in the quark covariant derivative
    \begin{equation}\label{eqn:covariant_deriv}
        (D_{\mu})_{ij} = \delta_{ij}\partial_{\mu} - ig_{s}T^{a}_{ij}A^{a}_{\mu} \, ,
    \end{equation}
    and the gauge field strength tensor
    \begin{equation}\label{eqn:F_munu}
        F^{a}_{\mu\nu} = \partial_{\mu}A^{a}_{\nu} - \partial_{\nu}A^{a}_{\mu} + g_{s} f^{abc}A^{b}_{\mu}A^{c}_{\nu} \, .
    \end{equation}
    Gauge fields transform under the adjoint representation, 
    which is an $N_{c}^{2}-1$ dimensional representation, meaning
    the adjoint indices $a,b,c \in \{1,...,8\}$. $T^{a}_{ij}$ are the
    group generators in a fundamental representation. In SU(3) it is
    common to write the group generators as
    \begin{equation}\label{eqn:group_generators}
        T^{a}_{ij} = \dfrac{1}{2}\lambda^{a}_{ij}
    \end{equation}
    where $\lambda^{a}_{ij}$ are the Gell-Mann matrices \cite{Gell-Mann:1962yej}.
    The generators of the group have to obey the Lie algebra
    \begin{equation}\label{eqn:lie_algebra}
        [T^{a}, T^{b}] = \mathrm{i}f^{abc}T^{c} \, ,
    \end{equation}
    where $f^{abc}$ are the structure constants of SU(3).
    The generators are normalised by convention to be
    \begin{equation}\label{eqn:generator_normalisation}
        \mathrm{Tr}(T^{a}T^{b}) = \delta^{ab}T_{R} \, , \quad T_{R} = \dfrac{1}{2} \, .
    \end{equation}
    This normalisation sets the values of the Casimirs
    of the group as
    \begin{align}\label{eqn:casimirs}
        \begin{split}
            T^{a}_{ij}T^{a}_{jk} &= \delta{ik}C_{F} \, , \quad C_{F} = \dfrac{N_{c}^2-1}{2N_{c}} \, , \\
            f^{abc}f^{abd} &= \delta^{cd}C_{A} \, , \quad C_{A} = N_{c} \, ,
        \end{split}
    \end{align}
    where $C_{F}=4/3$ and $C_{A}=3$ for QCD.
    They are collectively referred to as colour factors.

    The gauge coupling of the group $g_{s}$
    is a dimensionless free parameter of the theory.
    It is common to use the strong coupling constant
    instead, 
    \begin{equation}\label{eqn:alpha_s}
        \alpha_{s} = \dfrac{g_{s}^{2}}{4\pi} \, .
    \end{equation}
    The strong coupling constant, is in fact not constant, but
    is dependent on the energy scale of the process. This
    is due to the process of renormalisation which will be discussed
    in section ...
    The implications of this is that at collider
    experiments where collisions are extremely high energy,
    the strong coupling constant becomes small, a
    property known as asymptotic freedom. A consequence
    of this is that predictions made in QCD can be expressed
    in the form of perturbative expansions in $\alpha_{s}$.

@@

    \subsection{QCD Feynman rules}\label{sec:qcd_feynman}
    Now that we have written down the QCD Lagrangian,
    we can examine the terms to write down how the
    gauge bosons interact with the fermions. A convenient
    way to do this is with Feynman rules, which can be
    pieced together to form Feynman diagrams. We will see
    that Feynman diagrams are a simple, but powerful, tool
    to systematically build all the possible ways a process
    can occur. This will be discussed further in the
    scattering amplitude section.

    Expanding (\ref{eqn:L_QCD}) and extracting the terms
    that mix the gauge and fermions fields, we get an interaction Lagrangian
    \begin{equation}\label{L_int}
        \mathcal{L}_{\mathrm{int}} = g_{s}A^{a}_{\mu}\bar{\psi}_{i}^{f}\gamma^{\mu}T^{a}_{ij}\psi_{j}^{f}-g_{s}f^{abc}(\partial_{\mu}A^{a}_{\nu})A^{b,\,\mu}A^{c,\,\nu} -\dfrac{g_{s}^{2}}{4}f^{abc}A^{b}_{\mu}A^{c}_{\nu}f^{ade}A^{d,\,\mu}A^{e,\,\nu} \, ,
    \end{equation}
    which we can interpret as follows: the first term
    is an interaction between a gluon, a quark, and an
    anti-quark, the second term is a triple gluon
    self-interaction and the third term is a four gluon
    self-interaction. These mixing terms can be recast
    as Feynman rules as interaction vertices.

@@

    \begin{figure}
        \includegraphics[width=\linewidth]{introduction/qcd_feynman_rules.pdf}
        \caption{QCD Feynman rules.}
        \label{fig:qcd_feynman_rules}
    \end{figure}

    In Figure~\ref{fig:qcd_feynman_rules} we have
    collected the QCD Feynman rules.
    Note that we have only written down the vertices,
    propagators, and external lines for the physical gluons and quarks,
    ghosts also have associated Feynman rules but we do
    not collect those here. A full list of Feynman rules
    in the SM can be found at Ref. \cite{Romao:2012pq}.

    From these Feynman rules, we identify that each
    quark-antiquark-gluon and triple gluon vertices
    are associated with one power of $\alpha_{s}$,
    and the four gluon vertex is associated with
    two powers of $\alpha_{s}$. This relationship
    between vertices and $\alpha_{s}$ allows
    us to systematically build up Feynman diagrams
    that correspond to a fixed order in $\alpha_{s}$.
    This point will be elaborated on in Section
    \ref{sec:xs} where we discuss scattering amplitudes
    and how they are related to Feynman diagrams.

\section{Factorisation theorem}\label{sec:factorisation}
    At the LHC, collisions occur between two protons,
    which have constituent quarks and gluons, collectively
    named partons.
    The application of QCD to describe phenomenon
    in these proton-proton collisions rests on the use of
    the factorisation theorem which enables the
    separation of soft and hard energy scales.
    Within the proton, there are composite quarks
    and gluons that are constantly absorbed and
    emitted on a timescale inversely proportional
    to the mass of the proton (or hadron in general).
    The timescale of this fluctuation is much longer
    than the timescale of the interaction between
    a parton and a highly energetic probing parton
    from another proton. This is the scenario
    that occurs at collisions at the LHC.
    During the collision, or the hard scattering,
    the probe is able to interact with a
    parton that is effectively frozen.
    The probing parton knows nothing of the
    proton being probed, except for the fact
    that it collided with a parton carrying
    a fraction of the proton momentum. The
    distribution of partons within a proton
    is process independent and a fundamental
    property of the proton, meaning it can be
    separated from the actual scattering between
    the partons.

@@

    For a review of PDFs and how they are determined
    see Ref. \cite{Ethier:2020way,Jimenez-Delgado:2013sma}.
    In practice, these PDF fits are accessed through the
    LHAPDF interface \cite{Buckley:2014ana} which provides
    PDF sets from multiple working groups \cite{PDF4LHCWorkingGroup:2022cjn,Hou:2019efy,Bailey:2020ooq,NNPDF:2021njg}.
    The evolution of PDFs between factorisation scales is possible
    through the use of the Dokshitzer-Gribov-Lipatov-Altarelli-Parisi (DGLAP)
    equations \cite{Dokshitzer:1977sg,Gribov:1972ri,Lipatov:1974qm,Altarelli:1977zs}.
    \begin{equation}\label{eqn:DGLAP}
        \mu_{F}^{2} \dfrac{\partial}{\partial \mu_{F}^{2}}
        \begin{pmatrix}
            f_{q/h}(x,\mu_{F}) \\
            f_{g/h}(x,\mu_{F})
        \end{pmatrix} \\
        = \dfrac{\alpha_{s}(\mu_{F})}{2\pi} \int_{x}^{1} \dfrac{\mathrm{d}z}{z}
        \begin{pmatrix}
            P_{qq}(\frac{x}{z}) & P_{qg}(\frac{x}{z}) \\
            P_{gq}(\frac{x}{z}) & P_{gg}(\frac{x}{z})
        \end{pmatrix}
        \begin{pmatrix}
            f_{q/h}(x,\mu_{F}) \\
            f_{g/h}(x,\mu_{F})
        \end{pmatrix} \, ,
    \end{equation}
    where $P_{ba}(\frac{x}{z})$ are splitting functions
    representing parton $b$ emitting a parton $a$ carrying
    a momentum fraction $z$. These splitting functions are calculable
    as a power series in $\alpha_{s}$ where they have been computed
    up to three-loops \cite{Moch:2004pa,Vogt:2004mw}.
    At leading order \cite{Altarelli:1977zs} they are
    \begin{align}\label{eqn:AP_kernels}
        P^{(0)}_{qq}(z) &= C_{F}\left[\dfrac{1+z^{2}}{(1-z)_{+}} + \dfrac{3}{2}\delta(1-z)\right] \\
        P^{(0)}_{qg}(z) &= T_{R}\left[z^{2} + (1-z)^{2}\right] \\
        P^{(0)}_{gq}(z) &= C_{F}\left[\dfrac{1+(1-z)^{2}}{z^{2}}\right] \\
        P^{(0)}_{gg}(z) &= 2C_{A}\left[\dfrac{z}{(1-z)_{+}}+\dfrac{1-z}{z}+z(1-z)\right] + \left(\dfrac{11C_{A}-4T_{R}n_{f}}{6}\right)\delta(1-z)
    \end{align}
    where the `$+$'-prescription has been used to regulate divergences
    at $z = 1$.
    \begin{equation}\label{eqn:plus_prescription}
        \int_{0}^{1} \mathrm{d}z \; [g(z)]_{+}f(z) = \int_{0}^{1}\mathrm{d}z \; g(z)(f(z)-f(1)) \, ,
    \end{equation}
    where the function $f(z)$ cancels the divergence if it is
    sufficiently smooth at $z=1$.
    At leading order, there is a physical interpretation
    of the splitting kernels as the probability of finding a parton
    $a$ in parton $b$ carrying a fraction $x$ of the parent parton.

    The momenta fractions, $x$, also link the
    squared centre-of-mass energies of the hadronic collision,
    denoted as $s$, to the partonic equivalent as
    \begin{equation}\label{eqn:E_cm_partonic}
        \hat{s} = x_{a}x_{b}s \, .
    \end{equation}

    $\hat{\sigma}_{ab \rightarrow n}(Q, \mu_{F}, \mu_{R})$
    is the partonic cross-section which descibes the interaction of
    partons $a$ and $b$ scattering into $n$ particles,
    where the scattering occurs at energy scale $Q$,
    which is often taken to the $\hat{s}$.
    Notice that the partonic cross-section depends on
    both the factorisation scale, $\mu_{F}$ and
    the renormalisation scale, $\mu_{R}$
    (see Section~\ref{sec:renormalisation}). This will
    be discussed in Section~\ref{sec:scale_variations}
    in the context of theoretical uncertainties.

@@

    The differential partonic cross-section can be written as
    \begin{equation}\label{eqn:dsigma}
        \mathrm{d}\hat{\sigma} = \dfrac{1}{\mathcal{F}}|\mathcal{M}_{2 \rightarrow n}|^{2} \mathrm{d}\Phi_{n} \, .
    \end{equation}
    $\mathcal{F}$, the flux factor is
    \begin{equation}\label{eqn:flux}
        \mathcal{F} = 4\sqrt{(p_{a}p_{b})^{2} - m_{a}^{2}m_{b}^{2}} \, ,
    \end{equation}
    which in the massless limit reduces to
    \begin{equation}\label{eqn:massless_flux}
        \mathcal{F} = 2(p_{a} + p_{b})^{2} = 2\hat{s} \, .
    \end{equation}
    It is common to use the massless limit because
    the centre-of-mass energy of a collider
    experiment is much greater than the mass
    of the colliding particles.

    $\mathrm{d}\Phi_{n}$ is the Lorentz-invariant
    phase-space which contains all the possible
    configurations of the $n$-particle final state.
    Absorbing the momentum conserving $\delta$-function
    from (\ref{eqn:matrix_element}),
    we can write it as
    % \begin{equation}\label{eqn:dlips}
    %     \mathrm{d}\Phi_{n} = (2\pi)^{d}\delta^{(d)}\left(p_{a} + p_{b} - \sum_{f} p_{f}^{\mu}\right) \prod_{f=1}^{n} \dfrac{\mathrm{d}^{d}p_{f}}{(2\pi)^{d-1}}\delta(p_{f}^{2} - m_{f}^{2})\theta(E_{f}) \, ,
    % \end{equation}
    \begin{equation}\label{eqn:dlips_4d}
        \mathrm{d}\Phi_{n} = (2\pi)^{4}\delta^{4}\left(p_{a} + p_{b} - \sum_{f=1}^{n} p_{f}^{\mu}\right) \prod_{f=1}^{n} \dfrac{\mathrm{d}^{4}p_{f}}{(2\pi)^{3}}\delta(p_{f}^{2} - m_{f}^{2})\theta(E_{f}) \, ,
    \end{equation}
    where the second $\delta$-function restricts
    the final state particles to be on mass-shell
    (on-shell). The step function selects only
    the positive energy solution for this constraint.
    This gives an intuitive picture of the partonic
    cross-section as the probability of a $2\rightarrow n$
    process occurring, summed over all the possible
    valid final-state configurations.

>>>>

    The SM is a gauge quantum field theory (QFT) where particles
    are described as excitations in quantum fields. It can be specified
    by the gauge group
    \begin{equation}\label{eqn:SM_gauge}
        \mathrm{SU}(3)_{c} \times \mathrm{SU}(2)_{L} \times \mathrm{U}(1)_{Y}\, ,
    \end{equation}
    where subscripts denote the charges of the gauge groups. The first gauge
    group with colour charge $c$ describes the interactions of the
    strong force within the theory of quantum chromodynamics (QCD), which we
    will elaborate more on in Section~\ref{sec:qcd}. The second
    and third gauge groups represent the electroweak sector of the SM with
    charges $L$ for left and $Y$ for hypercharge. Under electroweak
    spontaneous symmetry breaking (EWSB), this product becomes
    \begin{equation}\label{eqn:SM_SSB}
        \mathrm{SU}(2)_{L} \times \mathrm{U}(1)_{Y} \xrightarrow{\mathrm{EWSB}} \mathrm{U}(1)_{\mathrm{EM}} \, ,
    \end{equation}
    giving rise to the electromagnetic and weak forces we observe.
    The effect of gravity is considered to be negligible on the scales
    considered in the SM and so it is not described.
    Each of the three fundamental forces described by the SM is mediated
    by the exchange of a gauge boson. The gauge bosons mediating the strong
    and EM force are the massless gluon $g$ and massless photon $\gamma$,
    respectively. For the weak force the gauge bosons are the $W^{\pm}$
    and $Z^{0}$ bosons which attain a mass through the Higgs mechanism \cite{Englert:1964et,Higgs:1964pj,Guralnik:1964eu}
    during EWSB, which elucidates the Higgs boson $H$.

@@

\section{Introduction to quantum chromodynamics}\label{sec:qcd}
    \subsection{The QCD Lagrangian}\label{sec:L_qcd}
    Quantum chromodynamics is the sector of the SM that
    describes the strong interaction. 
    QCD is a non-Abelian gauge theory with gauge group
    $\mathrm{SU}(N_{c} = 3)_{c}$ where the charge is named colour.
    The gauge and fermion part of the QCD Lagrangian is
    \begin{equation}\label{eqn:L_QCD}
        \mathcal{L}_{\mathrm{QCD}} = -\dfrac{1}{4}F^{a}_{\mu\nu}F^{a, \, \mu\nu}
        + \sum_{f} \bar{\psi}_{i}^{f}(\mathrm{i}\slashed{D}_{ij} - \delta_{ij}m_{f})\psi_{j}^{f} \, ,
    \end{equation}
    where twice repeated indices are summed over.
    The fields $\psi_{i}^{f}$ ($\bar{\psi}_{i}^{f}$) are the fermions (antifermions),
    representing quarks (antiquarks) with flavours $f$: 
    up $u$, down $d$, strange $s$, charm $c$, top $t$, and bottom $b$,
    with masses $m_{f}$. They transform under the fundamental
    representation with indices $i, j \in \{1, 2, 3\}$, named colour indices.
    The gauge fields $A^{a}_{\mu}$, corresponding to gluons, 
    appear in the quark covariant derivative
    \begin{equation}\label{eqn:covariant_deriv}
        (D_{\mu})_{ij} = \delta_{ij}\partial_{\mu} - ig_{s}T^{a}_{ij}A^{a}_{\mu} \, ,
    \end{equation}
    and the gauge field strength tensor
    \begin{equation}\label{eqn:F_munu}
        F^{a}_{\mu\nu} = \partial_{\mu}A^{a}_{\nu} - \partial_{\nu}A^{a}_{\mu} + g_{s} f^{abc}A^{b}_{\mu}A^{c}_{\nu} \, .
    \end{equation}
    Gauge fields transform under the adjoint representation, 
    which is an $N_{c}^{2}-1$ dimensional representation, meaning
    the adjoint indices $a,b,c \in \{1,...,8\}$. $T^{a}_{ij}$ are the
    group generators in a fundamental representation. In SU(3) it is
    common to write the group generators as
    \begin{equation}\label{eqn:group_generators}
        T^{a}_{ij} = \dfrac{1}{2}\lambda^{a}_{ij}
    \end{equation}
    where $\lambda^{a}_{ij}$ are the Gell-Mann matrices \cite{Gell-Mann:1962yej}.
    The generators of the group have to obey the Lie algebra
    \begin{equation}\label{eqn:lie_algebra}
        [T^{a}, T^{b}] = \mathrm{i}f^{abc}T^{c} \, ,
    \end{equation}
    where $f^{abc}$ are the structure constants of SU(3).
    The generators are normalised by convention to be
    \begin{equation}\label{eqn:generator_normalisation}
        \mathrm{Tr}(T^{a}T^{b}) = \delta^{ab}T_{R} \, , \quad T_{R} = \dfrac{1}{2} \, .
    \end{equation}
    This normalisation sets the values of the Casimirs
    of the group as
    \begin{align}\label{eqn:casimirs}
        \begin{split}
            T^{a}_{ij}T^{a}_{jk} &= \delta_{ik}C_{F} \, , \quad C_{F} = \dfrac{N_{c}^2-1}{2N_{c}} \, , \\
            f^{abc}f^{abd} &= \delta^{cd}C_{A} \, , \quad C_{A} = N_{c} \, ,
        \end{split}
    \end{align}
    where $C_{F}=4/3$ and $C_{A}=3$ for QCD.
    They are collectively referred to as colour factors.

    The gauge coupling of the group $g_{s}$
    is a dimensionless free parameter of the theory.
    It is common to use the strong coupling constant
    instead, 
    \begin{equation}\label{eqn:alpha_s}
        \alpha_{s} = \dfrac{g_{s}^{2}}{4\pi} \, .
    \end{equation}
    The strong coupling constant, is in fact not constant, but
    is dependent on the energy scale of the process. This
    is due to the process of renormalisation which will be discussed
    in Section~\ref{sec:renormalisation}.
    The implication of this is that at collider
    experiments where collisions occur at extremely high energy,
    the strong coupling constant becomes small, a
    property known as asymptotic freedom. A consequence
    of this is that predictions made in QCD can be expressed
    in the form of perturbative expansions in $\alpha_{s}$.

@@

    \subsection{QCD Feynman rules}\label{sec:qcd_feynman}
    Now that we have written down the QCD Lagrangian,
    we can examine the terms to write down how the
    gauge bosons interact with the fermions, and especially
    important for QCD, the gauge boson self interactions. A convenient
    way to do this is by deriving the Feynman rules of the theory, which can be
    pieced together to form Feynman diagrams. We will see
    that Feynman diagrams are a simple but powerful tool
    to systematically build up all the possible ways a process
    can occur.

    Expanding (\ref{eqn:L_QCD}) and extracting the terms
    that mix the gauge and fermions fields, we get an interaction Lagrangian
    \begin{equation}\label{L_int}
        \mathcal{L}_{\mathrm{int}} = g_{s}A^{a}_{\mu}\bar{\psi}_{i}^{f}\gamma^{\mu}T^{a}_{ij}\psi_{j}^{f}-g_{s}f^{abc}(\partial_{\mu}A^{a}_{\nu})A^{b,\,\mu}A^{c,\,\nu} -\dfrac{g_{s}^{2}}{4}f^{abc}A^{b}_{\mu}A^{c}_{\nu}f^{ade}A^{d,\,\mu}A^{e,\,\nu} \, ,
    \end{equation}
    which we can interpret as follows: the first term
    is an interaction between a gluon, a quark, and an
    anti-quark, the second term is a triple gluon
    self-interaction, and the third term is a four gluon
    self-interaction. These mixing terms can be recast
    into Feynman rules representing interaction vertices.

@@

    \begin{figure}
        \includegraphics[width=\linewidth]{introduction/qcd_feynman_rules.pdf}
        \caption{QCD Feynman rules with the Minkowski metric signature $g^{\mu\nu} = \mathrm{diag}(1, -1, -1, -1)$.}
        \label{fig:qcd_feynman_rules}
    \end{figure}

    From these Lagrangians, we can read off the QCD Feynman rules
    which we have collected in Figure~\ref{fig:qcd_feynman_rules}.
    Note that we have only written down the vertices and propagators
    for the physical gluons and quarks,
    ghosts also have associated Feynman rules but we do
    not collect those here. A full list of Feynman rules
    in the SM can be found at Ref. \cite{Romao:2012pq}.

    From these Feynman rules, we identify that each
    quark-antiquark-gluon and triple gluon vertices
    are associated with one power of $\alpha_{s}$ (due to squaring $g_{s}$),
    and the four gluon vertex is associated with
    two powers of $\alpha_{s}$. This relationship
    between vertices and $\alpha_{s}$ allows
    us to systematically build up Feynman diagrams
    that correspond to a fixed order in $\alpha_{s}$.
    This point will be elaborated on in Section
    \ref{sec:xs} where we discuss scattering amplitudes
    and how they are related to Feynman diagrams.

\section{Factorisation theorem}\label{sec:factorisation}
    At the LHC collisions occur between two protons,
    which have constituent quarks and gluons, collectively
    named partons.
    The application of QCD to describe phenomenon
    in these proton-proton collisions rests on the use of
    the factorisation theorem which enables the
    separation of low and high energy scales (so-called soft and hard).
    Within the proton, there are quarks
    and gluons that are constantly absorbed and
    emitted on a timescale inversely proportional
    to the mass of the proton.
    The timescale of this fluctuation is much longer
    than the timescale of the interaction between
    a parton and a highly energetic probing parton
    from another proton. This is the scenario
    that occurs at collisions at the LHC.
    During the collision, or the hard scattering,
    the probe is able to interact with a
    parton that is effectively frozen.
    The probing parton knows nothing of the
    proton being probed, except for the fact
    that it collided with a parton carrying
    a fraction of the proton momentum. The
    distribution of partons within a proton
    is process independent and a fundamental
    property of the proton, meaning it can be
    separated from the actual scattering between
    the partons.

@@

    For a review of PDFs and how they are determined
    see Refs. \cite{Ethier:2020way,Jimenez-Delgado:2013sma}.
    In practice, these PDF fits are accessed through the
    LHAPDF interface \cite{Buckley:2014ana} which provides
    PDF sets from multiple working groups \cite{PDF4LHCWorkingGroup:2022cjn,Hou:2019efy,Bailey:2020ooq,NNPDF:2021njg}.
    The evolution of PDFs between factorisation scales is possible
    through the use of the Dokshitzer-Gribov-Lipatov-Altarelli-Parisi (DGLAP)
    equations \cite{Dokshitzer:1977sg,Gribov:1972ri,Lipatov:1974qm,Altarelli:1977zs}.
    \begin{equation}\label{eqn:DGLAP}
        \mu_{F}^{2} \dfrac{\partial}{\partial \mu_{F}^{2}}
        \begin{pmatrix}
            f_{q/h}(x,\mu_{F}) \\
            f_{g/h}(x,\mu_{F})
        \end{pmatrix} \\
        = \dfrac{\alpha_{s}(\mu_{F})}{2\pi} \int_{x}^{1} \dfrac{\mathrm{d}z}{z}
        \begin{pmatrix}
            P_{qq}(\frac{x}{z}) & P_{qg}(\frac{x}{z}) \\
            P_{gq}(\frac{x}{z}) & P_{gg}(\frac{x}{z})
        \end{pmatrix}
        \begin{pmatrix}
            f_{q/h}(x,\mu_{F}) \\
            f_{g/h}(x,\mu_{F})
        \end{pmatrix} \, ,
    \end{equation}
    where $P_{ba}(\frac{x}{z})$ are splitting functions
    representing parton $b$ emitting a parton $a$ carrying
    a momentum fraction $x$. These splitting functions are calculable
    as a power series in $\alpha_{s}$ where they have been computed
    up to three-loops \cite{Moch:2004pa,Vogt:2004mw}.
    At leading order \cite{Altarelli:1977zs} they are
    \begin{align}\label{eqn:AP_kernels}
        P^{(0)}_{qq}(z) &= C_{F}\left[\dfrac{1+z^{2}}{(1-z)_{+}} + \dfrac{3}{2}\delta(1-z)\right] \\
        P^{(0)}_{qg}(z) &= T_{R}\left[z^{2} + (1-z)^{2}\right] \\
        P^{(0)}_{gq}(z) &= C_{F}\left[\dfrac{1+(1-z)^{2}}{z^{2}}\right] \\
        P^{(0)}_{gg}(z) &= 2C_{A}\left[\dfrac{z}{(1-z)_{+}}+\dfrac{1-z}{z}+z(1-z)\right] + \left(\dfrac{11C_{A}-4T_{R}n_{f}}{6}\right)\delta(1-z)
    \end{align}
    where the `$+$'-prescription has been used to regulate divergences
    at $z = 1$.
    \begin{equation}\label{eqn:plus_prescription}
        \int_{0}^{1} \mathrm{d}z \; [g(z)]_{+}f(z) = \int_{0}^{1}\mathrm{d}z \; g(z)(f(z)-f(1)) \, ,
    \end{equation}
    where the function $f(z)$ cancels the divergence if it is
    sufficiently smooth at $z=1$.
    At leading order, there is a physical interpretation
    of the splitting kernels as the probability of finding a parton
    $a$ in parton $b$ carrying a fraction $x$ of the parent parton.

    The momenta fractions, $x$, also link the
    squared centre-of-mass energies of the hadronic collision,
    denoted as $s$, to the partonic equivalent as
    \begin{equation}\label{eqn:E_cm_partonic}
        \hat{s} = x_{a}x_{b}s \, .
    \end{equation}
    $\hat{\sigma}_{ab \rightarrow n}(Q, \mu_{F}, \mu_{R})$
    is the partonic cross-section which descibes the interaction of
    partons $a$ and $b$ scattering into $n$ particles,
    where the scattering occurs at energy scale $Q$,
    which is often taken to be $\hat{s}$.
    Notice that the partonic cross-section depends on
    both the factorisation scale, $\mu_{F}$ and
    the renormalisation scale, $\mu_{R}$
    (see Section~\ref{sec:renormalisation}). This will
    be discussed in Section~\ref{sec:scale_variations}
    in the context of theoretical uncertainties.

@@

    The differential partonic cross-section can be written as
    \begin{equation}\label{eqn:dsigma}
        \mathrm{d}\hat{\sigma} = \dfrac{1}{\mathcal{F}}|\mathcal{M}_{2 \rightarrow n}|^{2} \mathrm{d}\Phi_{n} \, .
    \end{equation}
    $\mathcal{F}$, the flux factor is
    \begin{equation}\label{eqn:flux}
        \mathcal{F} = 4\sqrt{(p_{a}p_{b})^{2} - m_{a}^{2}m_{b}^{2}} \, ,
    \end{equation}
    which in the massless limit reduces to
    \begin{equation}\label{eqn:massless_flux}
        \mathcal{F} = 2(p_{a} + p_{b})^{2} = 2\hat{s} \, .
    \end{equation}
    It is common to use the massless limit because
    the centre-of-mass energy of a collider
    experiment is much greater than the mass
    of the colliding particles, and so they can
    be neglected.

    $\mathrm{d}\Phi_{n}$ is the Lorentz-invariant
    phase-space which contains all the possible
    configurations of the $n$-particle final-state.
    Absorbing the momentum conserving $\delta$-function
    from (\ref{eqn:matrix_element}),
    we can write it as
    % \begin{equation}\label{eqn:dlips}
    %     \mathrm{d}\Phi_{n} = (2\pi)^{d}\delta^{(d)}\left(p_{a} + p_{b} - \sum_{f} p_{f}^{\mu}\right) \prod_{f=1}^{n} \dfrac{\mathrm{d}^{d}p_{f}}{(2\pi)^{d-1}}\delta(p_{f}^{2} - m_{f}^{2})\theta(E_{f}) \, ,
    % \end{equation}
    \begin{equation}\label{eqn:dlips_4d}
        \mathrm{d}\Phi_{n} = (2\pi)^{4}\delta^{4}\left(p_{a} + p_{b} - \sum_{f=1}^{n} p_{f}^{\mu}\right) \prod_{f=1}^{n} \dfrac{\mathrm{d}^{4}p_{f}}{(2\pi)^{3}}\delta(p_{f}^{2} - m_{f}^{2})\theta(E_{f}) \, ,
    \end{equation}
    where the second $\delta$-function restricts
    the final-state particles to be on mass-shell
    (on-shell). The step function selects only
    the positive energy solution.
    This gives an intuitive picture of the partonic
    cross-section as the probability of a $2\rightarrow n$
    process occurring, summed over all the possible
    valid final-state configurations.

    From the definition of the transition probability
    (\ref{eqn:S_prob}), it becomes clear
    that the matrix element is the object which
    relates back to the Lagrangian of the theory
    as it encodes the interactions of the particles.
    We saw in Section \ref{sec:qcd_feynman} that
    interactions could be codified into Feynman
    rules which are pieced together to construct
    Feynman diagrams. In this picture, matrix elements
    are exactly the sum over all Feynman diagrams
    \footnote{Since we are working on the level
    of $|\mathcal{M}|^{2}$, it would technically
    be the sum of all interference of Feynman diagrams.}
    for a particular process. However, there are
    an infinite number of Feynman diagrams you can
    draw as there is no constraints on how the
    vertices and propagators can be connected, as
    long as the diagram contains the desired initial
    and final-states.
    Fortunately because $\alpha_{s}$
    is small in high energy collisions, we can expand
    the matrix element as a perturbative series in
    $\alpha_{s}$ to write
    \begin{equation}\label{eqn:matrix_element}
        |\mathcal{M}|^{2} =  \alpha_{s}^{m} |\mathcal{M}|^{2}_{\mathrm{LO}} + \alpha_{s}^{m+1} |\mathcal{M}|^{2}_{\mathrm{NLO}} + \alpha_{s}^{m+2} |\mathcal{M}|^{2}_{\mathrm{NNLO}} + \mathcal{O}(\alpha_{s}^{m+3}) \, ,
    \end{equation}
    where $m$ corresponds to the powers of $\alpha_{s}$
    in the simplest Feynman diagrams you can draw
    for the process of interest.
    Since each term of this expansion is at a fixed-order
    in $\alpha_{s}$, it is possible to systematically
    compute the diagrams which contribute at the given order of $\alpha_{s}$.
    The set of diagrams that contribute at the lowest
    order in $\alpha_{s}$ are $|\mathcal{M}|^{2}_{\mathrm{LO}}$,
    where LO stands for leading order. The next term
    in the expansion corresponds to the leading order
    diagrams with an additional loop, or external leg,
    which contributes an extra factor of $\alpha_{s}$.
    This term is dubbed NLO for next-to-leading order
    in $\alpha_{s}$, and the following term
    next-to-next-to-leading order has again an additional
    loop or leg. In general, we have terms
    N$^{\mathrm{k}}$LO where each additional power of $\alpha_{s}$
    increases complixity of the computations, however,
    because the expansion $\alpha_{s}$
    is small, each additional term should contribute
    less and less.
    This means that the first terms dominate the expansion,
    and it is sufficient to terminate the series after
    a few terms to reach an acceptable level of accuracy.
    A further discussion of this point will be given
    in Chapter~\ref{chapter:qcd} where we break down
    the computation of each term in (\ref{eqn:matrix_element}).

    In summary, the partonic cross-section of a collision
    reduces to the computation of the matrix elements
    up to a fixed-order in the coupling parameter,
    which is then integrated over the valid phase-space of
    final-state particle configurations. For hadronic
    collisions the partonic cross-section also needs to
    be convoluted with the PDFs to obtain the hadronic
    cross-sections.
